\chapter{Top Trees}

Top Trees are data structure intended to maintain informations of underlying
dynamically updated forest. They were introduced by Alstrup et al. in 2003 \cite{TopTrees}.

\section{Definition}

{\I Top Trees structure} acts as driver for underlying forest. It represents
underlying trees as collection of generalized edges called {\I clusters}. {\I
Clusters} are generalized edges and each of them represents some continuous
subtree in the underlying forest. Only some of them called {\I root clusters}
(which represents whole trees of the underlying forest) could be accessed by the
user.

User defines format of the data stored in these clusters and four basic
operations {\sc Create}, {\sc Destroy}, {\sc Join} and {\sc Split} used to
manipulate with clusters data.

Then user controls the Top Trees structure by using operations $\cut{u,v}$,
$\link{u,v}$ and $\expose{u,v}$. Last of them
makes cluster representing the path between vertices $u$ and $v$ a root cluster
(because root clusters are the only clusters of the top tree, whose could be
accessed by the user). The Top Trees structure dynamically updates stored data
in clusters by using user defined functions.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\section{Clusters}

As has been said {\I Clusters} are generalized edges. Each cluster has two
{\I boundary vertices} and represents part of the underlying forest between
these vertices.

Clusters in the Top Tree structure are connected into binary trees where each
leaf represents one edge of the underlying tree and each inner vertex represents
contraction of its children. More about this structure will be discussed
later in {\I Cluster model} subsection.

There are three types of clusters:

\begin{itemize}

\item {\bf Base cluster} -- represents one edge of the underlying tree (and each
edge of the underlying tree has exactly one base cluster, it is 1:1 mapping),
boundary vertices are endpoints of the edge.

This cluster could appear only as leaf of the top tree.

\item {\bf Rake cluster} -- represents one way how to contract two clusters
with one common boundary vertex. Lets have two clusters $C_1(u,v)$ and
$C_2(v,w)$ next to each other around common boundary vertex $v$ (and lets the
$C_1$ be the left one of them in some topological order given for example by
indexes of the edges).

Then we construct {\I left rake cluster} by {\I raking} the left cluster ($C_1$)
on the right one ($C_2$). The resulting cluster would have the same boundary
vertices as the cluster $C_2$.

Or we can construct {\I right rake cluster} in the same way -- by {\I raking}
the right cluster ($C_2$) on the left one ($C_1$). The resulting cluster would
have the same boundary vertices as the cluster $C_1$.

\todo{Image of rake cluster}

\item{\bf Compress cluster} -- represents other contraction of the two clusters
with one common boundary vertex into one cluster by attaching first cluster
after the other. If there are other clusters attached to the same common
boundary vertex they must be firstly {\I raked} onto one of the compressed
clusters.

If boundary vertices of the cluster $C_1$ were $(u,v)$ and boundary vertices
of the cluster $C_2$ were $(v,w)$, the cluster $C=compress(C_1,C_2)$ would have
boundary vertices $(u,w)$ (and we will call it sometime {\I compress cluster
of vertex $v$} or {\I handle of vertex $v$}).

\todo{Image of compress cluster}

\end{itemize}

\subsection{Clusters model}

Clusters in the Top Trees structure are organised into binary trees. Leaves of
these trees (Base clusters) represents edges of the underlying trees and inner
each inner vertex represents contraction of two adjacent clusters into one.

Compress and rake clusters have each of them two children, base clusters are
childless. Each cluster represent continuous subtree of the underlying forest.
By combination of clusters we could represent each underlying tree as one {\I
root cluster} (we will sometime refer to the representation of one continuous
underlying tree as one {\I top tree}).

Compress clusters are used to represent paths in the underlying tree -- each
path could be compressed into one {\I compress tree} consisting only of compress
clusters. If there are branches separating from this path, they are recursively
represented as single clusters ({\I rake trees}) and then they are
{\I raked onto} clusters in the path.

Because there are $N$ base clusters for an underlying tree with $N$ edges and
each inner vertex of the corresponding top tree joins two adjacent clusters into
one, there will be $N-1$ inner clusters for representing this underlying tree.

Underlying tree could have (and usually have) many different divisions into
paths and so the underlying tree have many different representations. Crucial
part of the top trees structure is to maintain this representation in some nice
form during updates.

\subsection{Extended clusters model}

Tarjan and Werneck in \cite{SelfAdjustingTT} suggested that in some cases it may
be useful to modify structure of the clusters and they introduced
{\I foster children} for {\I compress clusters}. In their suggestion a compress
cluster could have up to four descendants -- two normal children and up to two
foster children.

Normal children of a compress cluster are clusters from the compressed path and
foster children are clusters originating from the separating branches. In normal
cluster model they would be raked onto clusters from path and the path would be
compression of these rake clusters.

In this extended model the clusters originating from the separating branches are
firstly combined in so called {\I rake trees} -- there are maximally two rake
trees around each path vertex, one of them is raked from branches on one side of
the path and the second one is raked from branches on the other side of the
path. And these rake trees are connected as left and right foster child of the
compress cluster constructed from this part of the path.

\todo{Image of rake trees around path}

This extended model is used in the first implementation.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\section{User defined functions}

There are four basic function to manipulate with the clusters data which have to
be implemented by users of the Top Trees structure. They are called internally
by the public Top Trees operations when constructing, destroying or reorganizing
clusters.

If user want to use the {\I Search} operation, he has to implement fifth user
function {\I Choose} to traverse around the path by choosing children.

Examples of the functions and related problems are given in the Chapter 2.

\subsection{Create}

This function is called when new base cluster is created. It gets reference to
the underlying edge and to the newly created base cluster, populates base
clusters data based on the underlying edge and runs others user defined
operations according to logic of given problem.

\subsection{Destroy}

Opposite of the Create function. This function is called just before deleting
base cluster. It gets reference to the underlying edge and to the base cluster
which would be destroyed and it could perform some end-of-life operations (like
saving computed data from the cluster).

\subsection{Join}

This function is called when connecting two clusters into one (compress or rake)
cluster. It gets references to both of the connecting clusters and to the parent
cluster and it should aggregate data from child to the parent or perform other
join-related operations according to logic of given problem.

\subsection{Split}

Opposite of the Join function. It is called just before removing connection
between parent cluster and its children. This function gets reference to the
parent cluster and both of its children and it should distribute data from the
parent into children.

The Join and Split operations are frequently called during reorganization of the
top tree structure -- common pattern is to Split everything around changed path
in the top-down manner, reorganize the structure and then Join everything in the
bottom-up manner.

\subsection{Choose}

This operation for given cluster selects one of its child clusters. It is used
by the {\I Search} operation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\section{Top Trees operations}

These are the only operations which could user use to manipulate with the Top
Trees structure. In addition to that user could access root clusters and read
information from them.

\subsubsection{Handles}

Following operations are defined for vertices of the underlying forest, but the
Top Trees structure operates on (generalized) edges. To accomplish translation
of vertices onto edges we define {\I handle} for each vertex of the underlying
forest.

Isolated vertex has no handle. If the vertex is leaf of the underlying tree the
handle for this vertex is the topmost compress (or base) cluster having this
vertex as one of its endpoints. And finally if the vertex has degree at least
two the compress cluster of this vertex (compress cluster having this vertex
as the common boundary vertex) is the handle of this vertex.

\subsection{Expose}

This is the most basic operation (which is used internally by others
operations). Calling $\expose{u,v}$ will result into several changes in the Top
Trees structure depending on the positions of vertices $u$ and $v$.

Implementation of the expose slightly differ in the first and the second
implementation (first implementation uses splays and splices), but the basic
principle is common.

\subsubsection{Soft expose}

If they are in different components (they are not connected by path) both the
handles of $u$ and $v$ are brought to the roots of corresponding top trees
(similarly if $u=v$).

When they are in the same component (they are connected by path) firstly the
handle of $u$ is brought to the root of corresponding top tree. If the same
cluster is also the handle of $v$ we are done, otherwise the handle of $v$ is
brought as close to the root as possible (but not replacing the handle of $u$ as
root).

\subsubsection{Hard expose}

Ideal situation is when $u$ and $v$ are endpoints of the root cluster, it is
possible when both of them have degree one. But in general root cluster could
represent some path $(x,y)$ with path $(u,v)$ as subpath. In this case we need
to temporarily convert ends of this path (paths $(x,u)$ and $(v,y)$) into rake
clusters so the compress tree would represent the path $(u,v)$ with these ends
raked onto this path.

\subsection{Cut}

Operation $\cut{u,v}$ deletes edge between vertices $u$ and $v$ and reorganizes
the Top Trees structure to reflect this change. We assume that $u\ne v$ and
there is edge $(u,v)$.

Firstly the structure internally performs {\I soft expose} on $u,v$ to bring
theirs handles to the top of corresponding top tree. In this situation the
handle of $u$ will be the root cluster and the handle of $v$ will be child of
the root cluster.

We have to destroy the base node representing $(u,v)$ and remove connection
between these two handles, some reorganization of child clusters is needed (it
is implementation specific). This operation results in two top trees, one with
handle of~$u$ as root cluster and second with handle of~$v$ as root cluster.

\subsection{Link}

Calling $\link{u,v}$ on two disjoint vertices joins them by new edge $(u,v)$.
This is similar to the cut operation but in the opposite way. Firstly we start
by {\I exposing} (bringing to root) handles of $u$ and $v$.

Then we create new base cluster representing edge $(u,v)$, connect it on the
right place into one of the top trees, connect this tree as child of the root
cluster of the second tree and reorganize clusters (it is implementation
specific).

\subsection{Search}

When defined the {\I Choose} user function this operation could be used to find
and return specific cluster.
